\vspace{5cm} \normalsize \textsc{\NetIDb} \\ [2.0cm] \HRule{0.5pt} \\ \LARGE \textbf{\uppercase{\documenttitle}} \HRule{2pt} \\ [0.5cm] \normalsize \textsc{Sergio Castañeiras Morales\\NIU:1598456\\ }\normalsize \vspace*{5\baselineskip}

### 2025-05-02 Manuscript/Plantilla memòria TFG 2024_25.tex
writting

The target protein's ID is set at \emph{CHEMBL230} corresponding to the COX-2 ID in the ChEMBL database\cite{ChemblDatabase}. Utilising \emph{requests} python package\cite{PythonPackageRequests} a query URL is sent asking for all molecules with a know $IC_{50}$ value (Def. \ref{definitionIC50}) with a limit of 1000 entries per request. The process is iterated until all data is extracted leading a total of 7979 molecules. Hence the datasheet is processed in pandas dataframes\cite{PythonPackagePandas} and encrypted into binary feather files to optimise reading-writting speed. By removing entries with the same canonical smiles a total of 5112 molecules remain. Among this entries well known drugs such as Celecoxib (Def. \ref{definitionCelecoxib}), Rofecoxib (Def. \ref{definitionRofecoxib}) or even Ibuprofen can be found. However the $IC_{50}$ molecules range is comprehended from $10^{-3}$ to $10^{8}$ nM, a counterproductive range for the AI training procedure. Since we are interested in testing azophotoswitches with presumably low \gls{ic50}, training \gls{ml} models with data of the order of $10^{8}$ or $10^{3}$ nM can be counterproductive since the model might misinterpret the data.\footnote{Typically the range of the azophotoswitches \gls{ic50} is arround the Celecoxib's \gls{ic50}, i.e. arround $120$ nM}. Consecutivelly, a hard-coded range is filtered discarding all molecules outside the given range, for the most part of the analysis this range is set at $[0,200]$ nM \footnote{this $IC_{50}$ working range is the standard in this kind of studies \cite{MachineLearningPaper5Lipoxygenase}.} which reduces the dataset to 1438 entries (i.e. molecules).

With the AlvaDesk-python \cite{AlvaDescSecondPaper} facility, the chemical descriptors (i.e., the chemical fingerprint) of each molecule are computed, providing a total of $5800$ descriptors per molecule. Still, some chemical descriptor need from the presence of a certain atom, for instance a chemical descriptor related to presence of a metal in the molecule, or active gorup, for example chemical descriptor related to the presence of a carboxilic acid, or chemical descriptor related triple bound among other possiblities. The computation of this chemical descriptors appear as \emph{null values}. By deleting molecular descriptors with null values 2917 chemical descriptors remain. Here, the Pearson correlation coefficient (Def. (\ref{definitionPersonCoefficient})) between each chemical descriptor and the $IC_{50}$ value is computed, providing insight into the direct relationship between $IC_{50}$ and the descriptors. This relationship will be discussed in the Results and Discussion Section (\ref{sectionResultsAndDiscussion}).

@@

Subsequently, each set of substances is randomly divided into two datasets: a \emph{training set} for training the \gls{rf} algorithm and a \emph{testing set} for testing the statistics and realiability of the \gls{rf} model, following the proportion specified by the hard-coded variable \emph{testSizeProportion}. This procedure is illustrated in Figure (\ref{FigureDataSplittingDiagram}).
\begin{figure}[H]
\centering
\includegraphics[width = \textwidth]{GeneralSources/DataSplittingDiagram.pdf}
\caption{Splitting and processing data's scheme.}
\label{FigureDataSplittingDiagram}
\end{figure}
Afterwards, a Random Forest (\gls{rf}) algorithm is trained using the Sticky Learn \cite{PythonPackageStickitLearn} Python package, which is supported by Microsoft and Google among others. Although the project’s primary focus is not the detailed workings of Machine Learning (\gls{ml}) algorithms but its applications in computational chemestry, it is natural to ponder the essence of what the machine is doing. The name “Random Forest” is directly tied to the analogy between a tree and a decision tree. \par

As described in Definition (\ref{definitionDecisionTree}), the Random Forest algorithm generates a collection of random decision trees. The number of trees in the forest is a configurable parameter of the algorithm. More specifically, the Random Forest algorithm produces a set of decision trees, each trained on a random subset of the training data. The structure of each tree adjusts during training to make predictions that closely approximate the true target values. The final output of the Random Forest is typically an aggregate of the predictions from all the individual trees, such as a majority vote for classification tasks or an average for regression tasks. \par

One notable characteristic of this method is its inherent randomness. Each Random Forest model is unique, even if configured with the same parameters and initialized under identical conditions. However, experimental evidence suggests that this randomness does not significantly impact the results. If two Random Forest algorithms are trained on the same dataset using identical parameters, their outputs will converge, even if the internal structures of their decision trees differ.

The \gls{rf} algorithm generates a large number of random decision trees (as defined by the hard-coded variable \emph{numberOfTrees}), which are trained on subsets of the training data. Typically, increasing the number of trees enhances the model’s predictive accuracy; however, this comes at the cost of greater computational complexity. Importantly, prediction accuracy cannot be arbitrarily improved simply by increasing the number of trees. The model’s performance is fundamentally limited by the quality and quantity of the training data. If the data lacks sufficient information about the underlying protein-related phenomena, higher accuracy cannot be achieved regardless of how many trees are used.

In general, the relationship between data quality/quantity and the number of trees can be summarized as:
\begin{align*}
\uparrow\text{Data} + \uparrow\text{Trees} &\Rightarrow \uparrow\text{Prediction Accuracy} \\
\downarrow\text{Data} + \downarrow\text{Trees} &\Rightarrow \downarrow\text{Prediction Accuracy}
\end{align*}
However, increasing either component in isolation—more data or more trees—does not necessarily lead to better accuracy. For the available dataset in this work, we observed that training with 500 trees yields similar results to training with 300 trees.

@@

\subsection{Available Data}\label{sectionAvailableData}
Figure (\ref{figureGraphicRecurrecyIC50}) presents the \gls{ic50} values extracted from the ChEMBL database under target ID \emph{CHEMBL230}, limited to the interval $[0, 200]$ nM. The figure clearly shows that multiples of 10 are significantly overrepresented compared to other values. Upon analyzing data from additional proteins, we observe that this pattern is not unique to this target but appears to be a general trend. This suggests that experimental chemists tend to round their results to the nearest multiple of 10, introducing a systematic bias. As a result, a certain degree of inherent error will always be present in the dataset, and this limitation must be acknowledged and accounted for in subsequent analyses. This observation forms a key foundation for our later results.

@@

Figure (\ref{figureGraphicRecurrecyIC50}) also illustrates the process of generating a gap in the data. Given a dataset, the average \gls{ic50} is first computed—approximately 47 nM in this case. A proportion of the data, defined by the hard-coded variable \emph{percentageErased}, is then removed to create a gap centered around the average value. In the figure, the full dataset is represented in blue and orange, while the blue region, centered at 47 nM, has been deleted to introduce the gap. In this specific example, the gap accounts for 20\% of the total data.

@@

Next, chemical descriptors were calculated using the AlvaDesc software. Subsequently, we computed the Pearson Correlation Coefficients\footnote{The Pearson Correlation Coefficient is sometimes denoted as $r^2$, though strictly speaking $r^2$ refers to the coefficient of determination, which is the square of the correlation coefficient.} as defined in Definition (\ref{definitionPersonCoefficient}), measuring the linear relationship between each descriptor and the corresponding \gls{ic50} value for every molecule. The results are visualized in Figure (\ref{figureGraphicCorrelationFactor}).

@@

The distribution of correlation values resembles a Gaussian bell curve centered around zero, as expected. This suggests that there is no dominant or inherently preferred chemical descriptor strongly associated with \gls{ic50} variation for this protein. While this may seem to conflict with the project's objective, it actually reinforces the importance of using ensemble methods—such as \gls{rf}, which leverage the collective contribution of weak predictors.

This result is entirely expected. If a universally preferred descriptor existed, the use of \gls{ai} methods would be largely unnecessary, and significant advancements would likely have already been achieved through simpler statistical models. Nonetheless, our central hypothesis remains valid: \emph{there exists at least one combination of chemical descriptors that correlates directly with the inhibition of the target protein}. The observed distribution of correlation values does not contradict this hypothesis; rather, it highlights the need for models that can identify and exploit subtle multivariate relationships.

@@

Once the models have been generated, i.e., once the relevant hyperparameters are set, we proceed with prediction. At this stage, we compute the Pearson Correlation Coefficient (Def. (\ref{definitionPersonCoefficient})) and the Mean Squared Error (Def. (\ref{definitionMeanSquaredError})) between the predicted and experimental \gls{ic50} values.

@@

From the discussion in Section \ref{sectionAvailableData}, we understand that the dataset suffers from inherent precision issues, particularly at higher $IC_{50}$ values. Additionally, the precision of several measurements remains unknown. Since our models do not incorporate uncertainty estimates for each individual data point, it is unrealistic to expect $r^2$ and mean squared error values of exactly 1 and 0, respectively. Some degree of variance is inevitable due to the nature of the data. As such, discarding more than 50\% of the dataset, regardless of what statistical metrics may suggest, would be a critical error, as it would lead to substantial information loss. Still, the computed values of $r^2$ and mean squared error remain useful indicators of the model's performance up to a certain threshold.

@@

In this work, we arbitrarily define this threshold at $r^2 = 0.90$. This hard-coded value is a conventional choice, established to determine the point beyond which further improvements in $r^2$ become statistically irrelevant, given the underlying noise in the data. This convention is loosely inspired by standards in analytical chemistry,\footnote{Particularly in analytical chemistry, two datasets are typically considered significantly correlated if their $r^2$ value exceeds 0.90.} where correlations above 0.90 are often deemed acceptable. However, it is important to emphasize that this threshold is subjective and should not be treated as an absolute rule.

More reliable indicators of model performance are provided by the \emph{True Positive Rate} (Def. (\ref{definitionTruePositiveRate})), \emph{True Negative Rate} (Def. (\ref{definitionTrueNegativeRate})), \emph{Classification Accuracy} (Def. (\ref{definitionClassificationAccuracy})), and \emph{Matthews Correlation Coefficient} (Def. (\ref{definitionMatthewsCorrelationCoeficient})), the standard metrics used to evaluate the performance of a \gls{ml} model.

Moreover, we can hypothesize that creating a larger gap in the data facilitates the model’s ability to distinguish between highly active and weakly active inhibitors. This intuition is reflected in the boundary cases illustrated in Figure (\ref{FigPlotErasedPercentageVSRSquare}). As the erased percentage approaches 100\%, the $r^2$ value tends to zero and the mean squared error approaches one. This indicates that, in such extreme cases, the \gls{rf} model is able to classify only the most distinct molecules, those with \gls{ic50} values near the extremes of the interval. However, this is not a desirable scenario, as a significant portion of intermediate data is lost, limiting the model’s capacity to generalize across the full spectrum of inhibitor activity.

@@

Figure \ref{figureGraphicGbindingIC50} shows the predicted \gls{ic50} values of azophotoswitch compounds as a function of their binding free energy ($\Delta G_{binding}$), as provided by the Physical Chemistry Unit of the UAB. Detailed prediction results can be found in Appendix \ref{appendixTableOfResults}, specifically in Tables \ref{TableMainResults1} and \ref{TableMainResults2}. The training conditions for the Random Forest (\gls{rf}) model, along with a comprehensive statistical analysis of its predictions, are presented in Table~\ref{TableConditionsMainResults}.

@@

\begin{table}[H]
    \centering	
	\caption{Statistics for the computation of the results from tables (\ref{TableMainResults1}) and (\ref{TableMainResults2}).}
	\label{TableStatisitcsMainResults}
	\begin{tabular}{c|c}
		Mean Squared Error       & 1202.70\\\hline
		$r^2$ 	                 & 0.90\\\hline
		True Positive            & 994\\\hline
		False Positive           & 8\\\hline
		True Negative            & 1015\\\hline
		False Negative           & 101\\\hline
		True Positive Rate       & 0.90\\\hline
		True Negative Rate       & 0.99\\\hline
		ClassificationAcuracy    & 0.95\\\hline
		MatthewsCorrelationFactor& 0.90
    \end{tabular}
\end{table}
At this stage, it is important to note that the model is currently unable to distinguish between the active and inactive conformations of a molecule—specifically, between the \emph{cis} and \emph{trans} isomers. This limitation is not due to the model architecture or the way chemical descriptors are computed, but rather to a deficiency in the training data. The ChEMBL database \cite{ChemblDatabase} does not explicitly include \emph{cis}/\emph{trans} annotations for molecular configurations. Consequently, the model cannot learn to differentiate them if this distinction is not encoded in the dataset. As a result, the model provides \gls{ic50} predictions corresponding to the conformation it has implicitly learned, the more active configuration, which in our context is the \emph{cis} form.

Nevertheless, our hypothesis is supported. In Figure~\ref{figureGraphicGbindingIC50}, we observe a clear correlation between the predicted \gls{ic50} values and the corresponding binding free energies ($\Delta G_{binding}$), consistent with our initial intuition: molecules with more negative free energy values tend to exhibit lower \gls{ic50} values, and are therefore more active inhibitors.

The broader hypothesis—that a combination of chemical descriptors exists which correlates directly with the inhibitory power of a molecule toward a specific protein—also appears to be valid. However, identifying this combination manually remains a significant challenge for chemists. At first glance, one might believe that since this information is encoded within the \gls{rf} models, analyzing or "decrypting" the models could reveal valuable biochemical insights into the binding behavior of \gls{cox2},\footnote{Or any other protein, since the method is designed to be generalizable to any protein entry available in the ChEMBL database \cite{ChemblDatabase}.} or similar targets.

Yet this decryption process is computationally infeasible. The internal logic of a \gls{rf} model is akin to verifying a password: it can efficiently determine whether an input is valid (e.g., predict an \gls{ic50}), but reverse-engineering the logic—such as determining which chemical descriptors most strongly influence inhibition—is significantly more complex. In essence, it is easy for the model to make predictions, but very difficult to extract interpretable rules or patterns from its internal structure.

One might consider a brute-force approach: varying one descriptor at a time while holding all others constant, and observing the effect on the predicted \gls{ic50}. However, this strategy quickly runs into chemical and conceptual limitations. For example, altering the average bond length of a molecule could make it unrealistically large or physically unstable—resulting in a structure that is chemically invalid. Similarly, descriptors that are inherently discrete, such as the number of carbon atoms, could yield nonsensical results. A model might indicate that 25.5 carbon atoms yields optimal inhibition, which is clearly an unphysical outcome. In addition, arbitrarily generated sets of descriptor values may not correspond to any real molecule, making the predictions scientifically meaningless.

It is important to emphasize that these limitations do not undermine the value of \gls{ai} in computational chemistry. Rather, they highlight the boundaries of current methods and the importance of careful dataset design, descriptor selection, and model interpretation.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \textwidth, trim={0cm 0cm 0cm 0cm}, clip]{GeneralSources/LastFrameDockingTwoRingsFiveMembers.jpg}
	\caption{Structural ilustration of the molecule identied as \ref{figureCelecoxibTwoRings}.7 and \gls{cox2} docking generated by the ChimeraX software[15].}
	\label{figureDockingBestResult}
\end{figure}
We observe that the azophotoswitch in the \emph{cis} configuration fits well within the \gls{cox2} binding cavity. Although the \emph{trans} isomer can, in theory, occupy the binding site,\footnote{With minor steric hindrances.} it is unlikely to access it in practice due to stereochemical constraints. Specifically, the extended structure of the \emph{trans} configuration results in steric clashes with the outer regions of \gls{cox2}, preventing entry into the binding pocket.

It is also important to note that the molecule, in both \emph{E}/\emph{Z} isomeric forms, could potentially interact with unintended proteins. Such off-target effects could lead to significant side effects if the compound is administered as a drug. Therefore, further investigation is needed to evaluate its selectivity and pharmacological safety.

@@

Additionally, our results confirm the hypothesized correlation between predicted \gls{ic50} values and the experimentally or computationally derived binding free energy ($\Delta G_{binding}$). This relationship aligns with our chemical intuition and reinforces the validity of our methodology. To further validate our approach, we performed a molecular docking study of one of the most promising azophotoswitch candidates. The docking results were consistent with model predictions, providing additional support for the application of AI techniques in rational drug design.

>>>>

The target protein's ID is set at \emph{CHEMBL230} corresponding to the COX-2 ID in the ChEMBL database\cite{ChemblDatabase}. Utilising \emph{requests} python package\cite{PythonPackageRequests} a query URL is sent asking for all molecules with a know $IC_{50}$ value (Def. \ref{definitionIC50}) with a limit of 1000 entries per request. The process is iterated until all data is extracted leading a total of 7979 molecules. Hence the datasheet is processed in pandas dataframes\cite{PythonPackagePandas} and encrypted into binary feather files to optimise reading-writting speed. By removing entries with the same canonical smiles a total of 5112 molecules remain. Among this entries well known drugs such as Celecoxib (Def. \ref{definitionCelecoxib}), Rofecoxib (Def. \ref{definitionRofecoxib}) or even Ibuprofen can be found. However the $IC_{50}$ molecules range is comprehended from $10^{-3}$ to $10^{8}$ nM, a counterproductive range for the AI training procedure. Since we are interested in testing azophotoswitches with presumably low \gls{ic50}, training \gls{ml} models with data of the order of $10^{8}$ or $10^{3}$ nM can be counterproductive since the model might misinterpret the data.\footnote{Typically the range of the azophotoswitches \gls{ic50} is arround the Celecoxib's \gls{ic50}, i.e. arround $120$ nM}. Consecutivelly, a hard-coded range is filtered discarding all molecules outside the given range, for the most part of the analysis this range is set at $[0,300]$ nM \footnote{this $IC_{50}$ working range is the standard in this kind of studies \cite{MachineLearningPaper5Lipoxygenase}.} which reduces the dataset to 1438 entries, i.e. molecules.

With the AlvaDesk-python \cite{AlvaDescSecondPaper} facility, the chemical descriptors (i.e., the chemical fingerprint) of each molecule are computed, providing a total of $5800$ descriptors per molecule. Still, some chemical descriptor need from the presence of a certain atom, for instance a chemical descriptor related to presence of a metal in the molecule, or active gorup, for example chemical descriptor related to the presence of a carboxilic acid, or chemical descriptor related triple bound among other possiblities. The computation of this chemical descriptors appear as \emph{null values}. By deleting molecular descriptors with null values 2917 chemical descriptors remain. Here, the Pearson correlation coefficient (Def. \ref{definitionPersonCoefficient}) between each chemical descriptor and the $IC_{50}$ value is computed, providing insight into the direct relationship between $IC_{50}$ and the descriptors. This relationship will be discussed in the Results and Discussion Section (\ref{sectionResultsAndDiscussion}).

@@

Subsequently, each set of substances is randomly divided into two datasets: a \emph{training set} for training the \gls{rf} algorithm and a \emph{testing set} for testing the statistics and realiability of the \gls{rf} model, following the proportion specified by the hard-coded variable \emph{testSizeProportion}. This procedure is illustrated in Figure (\ref{FigureDataSplittingDiagram}).

\begin{figure}[H]
\centering
\includegraphics[width = \textwidth]{GeneralSources/DataSplittingDiagram.pdf}
\caption{Splitting and processing data's scheme.}
\label{FigureDataSplittingDiagram}
\end{figure}

Afterwards, a Random Forest (\gls{rf}) algorithm is trained using the Sticky Learn \cite{PythonPackageStickitLearn} Python package, which is supported by Microsoft and Google among others. Although the project’s primary focus is not the detailed workings of Machine Learning (\gls{ml}) algorithms but its applications in computational chemestry, it is natural to ponder the essence of what the machine is doing. The name “Random Forest” is directly tied to the analogy between a tree and a decision tree. \par

As described in definition (\ref{definitionDecisionTree}), the Random Forest algorithm generates a collection of random decision trees. The number of trees in the forest is a configurable parameter of the algorithm. More specifically, the Random Forest algorithm produces a set of decision trees, each trained on a random subset of the training data. The structure of each tree adjusts during training to make predictions that closely approximate the true target values. The final output of the Random Forest is typically an aggregate of the predictions from all the individual trees, such as a majority vote for classification tasks or an average for regression tasks. \par

One notable characteristic of this method is its inherent randomness. Each Random Forest model is unique, even if configured with the same parameters and initialized under identical conditions. However, experimental evidence suggests this randomness does not significantly impact the results. If two Random Forest algorithms are trained on the same dataset using identical parameters, their outputs will converge, even if the internal structures of their decision trees differ.

The \gls{rf} algorithm generates a large number of random decision trees (as defined by the hard-coded variable \emph{numberOfTrees}), which are trained on subsets of the training data. Typically, increasing the number of trees enhances the model’s predictive accuracy, however, this comes at the cost of greater computational complexity. Importantly, prediction accuracy cannot be arbitrarily improved simply by increasing the number of trees. The model’s performance is fundamentally limited by the quality and quantity of the training data. If the data lacks from information about the underlying protein-related phenomena, higher accuracy cannot be achieved regardless of how many trees are used.

In general, the relationship between data quality/quantity and the number of trees can be summarized as the following
\begin{align*}
\uparrow\text{Data} + \uparrow\text{Trees} &\Rightarrow \uparrow\text{Prediction Accuracy} \\
\downarrow\text{Data} + \downarrow\text{Trees} &\Rightarrow \downarrow\text{Prediction Accuracy}
\end{align*}
However, increasing either component in isolation, i.e. add more data or more trees, does not necessarily lead to better accuracy. For the available dataset in this work, we observed training with 500 trees yields similar results to training with 300 trees.

@@

\subsection{Available Data}\label{sectionAvailableData}
Figure (\ref{figureGraphicRecurrecyIC50}) presents the \gls{ic50} values extracted from the ChEMBL database under target ID \emph{CHEMBL230}, limited to the interval $[0, 200]$ nM. The figure clearly shows that multiples of 10 are significantly overrepresented compared to other values. Upon analyzing data from additional proteins, we can confirm this is not an exception but rather the norm. This suggests experimental chemists tend to round their results to the nearest multiple of 10, introducing a systematic bias. As a result, a certain degree of inherent error will always be present in the dataset, and this limitation must be acknowledged and accounted for in subsequent analyses. This observation forms a key foundation for our later results.

@@

Figure (\ref{figureGraphicRecurrecyIC50}) also illustrates the process of generating a gap in the data. Given a dataset, the average \gls{ic50} is first computed, approximately 47 nM in this case. A proportion of the data, defined by the hard-coded variable \emph{percentageErased}, is then removed to create a gap centered around the average value. In Figure (\ref{figureGraphicRecurrecyIC50}), the full dataset is represented in blue and orange, while the blue region, centered at 47 nM, has been deleted to introduce the gap. In this specific example, the gap accounts for 20\% of the total data.

@@

Afterwards, chemical descriptors are calculated using the AlvaDesc software. Subsequently, we compute the Pearson Correlation Coefficients\footnote{The Pearson Correlation Coefficient is sometimes denoted as $r^2$, though strictly speaking $r^2$ refers to the coefficient of determination, which is the square of the correlation coefficient.} as defined in Definition (\ref{definitionPersonCoefficient}), measuring the linear relationship between each descriptor and the corresponding \gls{ic50} value for every molecule. The results are visualized in Figure (\ref{figureGraphicCorrelationFactor}).

@@

The distribution of correlation values resembles a Gaussian bell curve centered around zero, as expected. This suggests that there is no dominant or inherently preferred chemical descriptor strongly associated with \gls{ic50} variation for this protein. While this may seem to conflict with the project's objective, it actually reinforces the importance of using ensemble methods.

This result is entirely expected. If a universally preferred descriptor existed, the use of \gls{ai} methods would be largely unnecessary, and significant advancements would likely have already been achieved through simpler statistical models. Nonetheless, our central hypothesis remains valid: "\emph{there exists at least one combination of chemical descriptors that correlates directly with the inhibition of the target protein}". The observed distribution of correlation values does not contradict this hypothesis, rather, it highlights the need for models that can identify and exploit subtle multivariate relationships.

@@

Once the models have been generated, i.e., once the relevant hyperparameters are set, we proceed with generating predictions. At this stage, we compute the Pearson Correlation Coefficient (Def. (\ref{definitionPersonCoefficient})) and the Mean Squared Error (Def. (\ref{definitionMeanSquaredError})) between the predicted and experimental \gls{ic50} values.

@@

From the discussion in Section \ref{sectionAvailableData}, we understand the dataset suffers from inherent precision issues, particularly at higher $IC_{50}$ values. Additionally, the precision of several measurements remains unknown. Since our models do not incorporate uncertainty estimates for each individual data point, it is unrealistic to expect $r^2$ and mean squared error values of exactly 1 and 0, respectively. Some degree of variance is inevitable due to the nature of the data. As such, discarding more than 50\% of the dataset, regardless of what statistical metrics may suggest, would be a critical error, as it would lead to substantial information loss. Still, the computed values of $r^2$ and mean squared error remain useful indicators of the model's performance up to a certain threshold.

@@

In this work, we arbitrarily define this threshold at $r^2 = 0.90$. This hard-coded value is a conventional choice, established to determine the point beyond which further improvements in $r^2$ become statistically irrelevant, given the underlying noise in the data. This convention is loosely inspired by standards in analytical chemistry,\footnote{Particularly in analytical chemistry, two datasets are typically considered significantly correlated if their $r^2$ value exceeds 0.90.} where correlations above 0.90 are often deemed acceptable. However, it is important to emphasize this threshold is subjective and should not be treated as an absolute rule.

Moreover, we can hypothesize creating a larger gap in the data facilitates the model’s ability to distinguish between highly active and weakly active inhibitors. This intuition is reflected in the boundary cases illustrated in Figure (\ref{FigPlotErasedPercentageVSRSquare}). As the erased percentage approaches 100\%, the $r^2$ value tends to zero and the mean squared error approaches one. This indicates that, in such extreme cases, the \gls{rf} model is able to classify only the most distinct molecules, those with \gls{ic50} values near the extremes of the interval. However, this is not a desirable scenario, as a significant portion of intermediate data is lost, limiting the model’s capacity to generalize across the full spectrum of inhibitor activity.

More reliable indicators of model's performance are provided by the \emph{True Positive Rate} (Def. \ref{definitionTruePositiveRate}), \emph{True Negative Rate} (Def. \ref{definitionTrueNegativeRate}), \emph{Classification Accuracy} (Def. \ref{definitionClassificationAccuracy}), and \emph{Matthews Correlation Coefficient} (Def. \ref{definitionMatthewsCorrelationCoeficient}), the standard metrics used to evaluate the performance of a \gls{ml} model. Generally speaking, we define a method A to have "better statisitcs" than other method B if the computed quantities: \emph{True Positive Rate}, \emph{True Negative Rate}, \emph{Classification Accuracy}, and \emph{Matthews Correlation Coefficient}, for A take higher values than the same quantities for B.\footnote{Since all quantities are highly bounded by 1 we might also affirm, the closer to 1 the statistics values are the more reliable the method is.} Specifically, if this procedure for the method A and B lead to incoherences, i.e. some statistics of A are better than B while other are worst, we take the Matthews Correlation Factor as the most reliable statistical outcome.\footnote{This is the usual procedure in the analysis of \gls{ml} predictions\cite{MachineLearningPaper5Lipoxygenase}.}

@@

Figure \ref{figureGraphicGbindingIC50} shows the predicted \gls{ic50} values of azophotoswitch compounds as a function of their binding free energy ($\Delta G_{binding}$), as provided by the Physical Chemistry Unit of the UAB \cite{UAB_ComputationalChemistry}. Detailed prediction results can be found in Appendix \ref{appendixTableOfResults}, specifically in Tables \ref{TableMainResults1} and \ref{TableMainResults2}. The training conditions for the Random Forest (\gls{rf}) model, along with a comprehensive statistical analysis of its predictions, are presented in Table~\ref{TableConditionsMainResults}.

@@

\begin{table}[H]
    \centering	
	\caption{Statistics for the computation of the results from tables (\ref{TableMainResults1}) and (\ref{TableMainResults2}).}
	\label{TableStatisitcsMainResults}
	\begin{tabular}{c|c}
		Mean Squared Error       & 1202.70\\\hline
		$r^2$ 	                 & 0.90\\\hline
		True Positive            & 994\\\hline
		False Positive           & 8\\\hline
		True Negative            & 1015\\\hline
		False Negative           & 101\\\hline
		True Positive Rate       & 0.90\\\hline
		True Negative Rate       & 0.99\\\hline
		ClassificationAcuracy    & 0.95\\\hline
		MatthewsCorrelationFactor& 0.90
    \end{tabular}
\end{table}
At this stage, it is important to note that the model is currently unable to distinguish between the active and inactive conformations of a molecule, i.e, between the \emph{cis} and \emph{trans} isomers. This limitation is not due to the model architecture or the way chemical descriptors are computed, but rather to a deficiency in the training data. The ChEMBL database \cite{ChemblDatabase} does not explicitly include \emph{cis}/\emph{trans} comparable molecular configurations. In other words, if (Z)/(E) isomerization appears we can only find one of the two conformations in the database, typically the one with higher inhibition potential. Consequently, the model cannot learn to differentiate them since this distinction is not encoded in the dataset. As a result, the model provides \gls{ic50} predictions corresponding to the conformation it has implicitly learned, the more active configuration, which in our context is the \emph{cis} form.

Nevertheless, our hypothesis is supported. In Figure~\ref{figureGraphicGbindingIC50}, we observe a clear correlation between the predicted \gls{ic50} values and the corresponding binding free energies ($\Delta G_{binding}$), consistent with our initial intuition that molecules with more negative free energy values tend to exhibit lower \gls{ic50} values, and are, therefore, more active inhibitors.

The broader hypothesis that a combination of chemical descriptors exists which correlates directly with the inhibitory power of a molecule toward a specific protein also appears to be valid. However, identifying this combination manually remains a significant challenge for chemists. At first glance, one might believe that since this information is encoded within the \gls{rf} models, analyzing or "decrypting" the models could reveal valuable biochemical insights into the binding behavior of \gls{cox2},\footnote{Or any other protein, since the method is designed to be generalizable to any protein entry available in the ChEMBL database \cite{ChemblDatabase}.} or similar targets.

Yet this decryption process is computationally infeasible. The internal logic of a \gls{rf} model is akin to verifying a password. It can efficiently determine whether an input is valid (e.g., predict an \gls{ic50} from a molecule's descriptors or checking a username and a password), but reverse-engineering (e.g., determining which chemical descriptors most strongly influence inhibition or cracking a user password) is significantly more complex. In essence, it is easy for the model to make predictions, but very difficult to extract interpretable rules or patterns from its internal structure.

One might consider a brute-force approach, varying one descriptor at a time while holding all others constant, and observing the effect on the predicted \gls{ic50}. However, this strategy quickly runs into chemical contradictions and incoherences. For example, altering the average bond length of a molecule could make it unrealistically large or physically unstable, resulting in a structure that is chemically invalid although this hypothetic molecule might perfectly fit in the \gls{cox2} binding site. Similarly, descriptors that are inherently discrete, such as the number of carbon atoms, could yield nonsensical results. A model might indicate that 25.5 carbon atoms yields optimal inhibition, which is clearly an unphysical outcome. In addition, arbitrarily generated sets of descriptor values may not correspond to any real molecule, making the predictions scientifically meaningless.

It is important to emphasize these limitations do not undermine the value of \gls{ai} in computational chemistry. Rather, they highlight the boundaries of current methods and the importance of careful dataset design, descriptor selection, model interpretation and understanding the machine's workflow.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \textwidth, trim={0cm 0cm 0cm 0cm}, clip]{GeneralSources/LastFrameDockingTwoRingsFiveMembers.jpg}
	\caption{Structural ilustration of the molecule identied as \ref{figureCelecoxibTwoRings}.7 and \gls{cox2} docking generated by the ChimeraX software[15].}
	\label{figureDockingBestResult}
\end{figure}
We observe the azophotoswitch in the \emph{cis} configuration happens to fit well within the \gls{cox2} binding cavity. Although the \emph{trans} isomer can, in theory, occupy the binding site,\footnote{With minor steric hindrances.} it is unlikely to access it in practice due to stereochemical constraints. Specifically, the extended structure of the \emph{trans} configuration results in steric clashes with the outer regions of \gls{cox2}, preventing entry into the binding pocket.

It is also important to note the molecule, in both \emph{E}/\emph{Z} isomeric forms, could potentially interact with unintended proteins. Such off-target effects could lead to significant side effects if the compound is administered as a drug. Therefore, further investigation is needed to evaluate its selectivity and pharmacological safety.

@@

Despite their usefulness, the models exhibit several limitations. They are entirely dependent on the training dataset, therefore, any lack of information or approximations in the data can lead to systematic errors in their predictions. Furthermore, the models are inherently unable to generate new knowledge. For instance, the distinction between cis and trans conformations is not represented in the ChEMBL \gls{cox2} database. As a result, our models are unable to provide different \gls{ic50} predictions for these two configurations. Nonetheless, we have demonstrated that the predictions are reliable within a certain threshold, as determined by the statistical analysis of each case.

Additionally, our results confirm the hypothesized correlation between predicted \gls{cox2}'s \gls{ic50} values and the experimentally or computationally derived binding free energy ($\Delta G_{binding}$). This relationship aligns with our chemical intuition and reinforces the validity of our methodology. To further validate our approach, we performed a molecular docking study of one of the most promising azophotoswitch candidates. The docking results were consistent with model predictions, providing additional support for the application of AI techniques in rational drug design.
